Expression QTL hotspot dissection with mediation analysis and pleiotropy testing

### 2019-08-12 analysis/paper/Rmd/methods.Rmd
revised first subsection of Methods

We used the publicly available data from @keller2018genetic to demonstrate our strategy. @keller2018genetic
measured genome-wide gene expression levels with RNA sequencing of pancreatic islets in 378 Diversity Outbred mice. From the RNA sequencing counts, they obtained normalized gene expression levels for 21,771 transcripts, which they shared publicly via DataDryad [@dryad]. @keller2018genetic also included founder allele probabilities at 69,005 markers across the genome, covariates, and leave-one-chromosome-out estimated kinship matrices. @keller2018genetic calculated leave-one-chromosome-out kinship matrices for every chromosome [@broman2019rqtl2; @yang2014advantages].

@keller2018genetic performed univariate QTL mapping for all expression traits. 
They identified five expression trait hotspots in their
study of pancreatic islet gene expression in 378 Diversity Outbred mice. They
defined a hotspot to be a 4-Mb region that affects at least 100 nonlocal expression
traits.

For each hotspot, we identified a set of local expression traits and a
set of nonlocal expression traits. We focused on those local 
traits that strongly associated with QTL within the hotspot. We set the univariate LOD threshold for 
local traits at 15. We limited consideration to those local traits for which the middle of the gene lay 
within the hotspot interval. 
All nonlocal traits demonstrated univariate LOD peaks that met
the genome-wide significance threshold, 7.18 [@keller2018genetic]. The nonlocal expression traits arose 
from genes
on other chromosomes, while the local traits reflected expression levels for
genes on the same chromosome as the hotspot . We ignored transcripts for genes on the hotspot chromosome 
but outside the hotspot interval.

>>>>

We used the publicly available data from @keller2018genetic to demonstrate our approach for 
expression trait hotspot dissection. @keller2018genetic measured genome-wide gene expression 
levels with RNA sequencing of pancreatic islets in 378 
Diversity Outbred mice. From the RNA sequencing counts, they obtained normalized gene 
expression levels for 21,771 transcripts, which they shared publicly via DataDryad [@dryad]. 
@keller2018genetic also included founder allele probabilities at 69,005 markers across the 
genome, covariates, and leave-one-chromosome-out estimated kinship matrices [@broman2019rqtl2; @yang2014advantages].

@keller2018genetic performed univariate QTL mapping for all expression traits. They
defined an expression trait hotspot to be a 4-Mb region that affects at least 100 nonlocal 
expression traits.
They identified five expression trait hotspots in their
study of pancreatic islet gene expression in 378 Diversity Outbred mice .

In analyzing the data from @keller2018genetic, we first established criteria for identifying local and nonlocal transcripts at each of the five hotspots. We focused on those local 
traits that strongly associated with QTL within the hotspot. We set the univariate LOD threshold for 
local traits at 15. We limited consideration to those local traits for which the middle of the gene lay 
within the hotspot interval. 
All nonlocal traits demonstrated univariate LOD peaks that met
the genome-wide significance threshold, 7.18 [@keller2018genetic]. The nonlocal expression traits arose 
from genes
on other chromosomes, while the local traits reflected expression levels for
genes on the same chromosome as the hotspot. We ignored transcripts for genes on the hotspot chromosome 
but outside the hotspot interval.

For each hotspot, we performed both pairwise pleiotropy tests and linear regression-based 
mediation analyses. For every pair of traits at a hotspot, we calculated a pleiotropy test 
statistic[@boehm2019testing]. Due to the computational burden, we didn't determine bootstrap p-values for these 
tests. For every local trait - nonlocal trait pair, we also performed a linear 
regression-based mediation analysis [@chick2016defining,@keller2018genetic]. We describe in detail both procedures below.
